FFT based convolution

### 2010-03-07 article/Article.tex
article continued
Ignore-this: 26996452fdf987ebfa8b4427885f5323

darcs-hash:20100307220037-2fc9d-cdc8c19d63e2e0382b190d594ee38cb974fcd5ba.gz

\subsection{FFTW filters enhancements}
\subsubsection{Thread support}
fft et norm
\subsubsection{Memory usage reduction}
\subsubsection{Code safety}

@@

@@

>>>>

\subsection{FFTW filters enhancements}

ITK provides FFT and inverse FFT filters based on the FFTW library. Those filters had some problems,
or were missing some features. They have been enhanced to fit the needs of the FFT based convolution.

\subsubsection{Thread support}
The FFTW filters were not multithreaded. There are two places where multithreading can take place: the
FFTW execution and the normalization step.

The FFTW execution can be threaded simply by calling \verb$fftwf_plan_with_nthreads(threads);$ or 
\verb$fftwf_plan_with_nthreads(threads);$, depending on the pixel type, before creating the plan. 
\verb$itk::fftw::Proxy$ has been modified to provide this feature: an extra \verb$thread$ parameter,
which defaults
to $1$ to keep the backward compatibility, has been added to the method in charge of creating the plans.
Then the \verb$itk::FFTWComplexConjugateToRealImageFilter$ and
\verb$itk::FFTWRealToComplexConjugateImageFilter$ have
been modified to pass their user defined number of threads to those methods.

In order to implement the threading support in the normalization step in 
\verb$itk::FFTWComplexConjugateToRealImageFilter$, the execution of the FFTW library has been moved to 
\verb$BeforeThreadedGenerateData()$, and the normalization step to \verb$ThreadedGenerateData()$.

\subsubsection{Memory usage reduction}

FFTW filters were using a lot of memory, because they were allocating an internal buffer for the input 
image, and an other one for the output image. Worth, these buffers were not deallocated at the end of
the execution filter, keeping the memory usage very high even when the execution has completed.

I guess the main reasons to do that were:
\begin{itemize}
  \item avoid the destruction of the input. The creation of the FFTW plan, and the execution of FFTW 
  can destroy the input. This is most of the time not acceptable in ITK.
  \item reuse the FFTW plan for a latter run. The plan include the memory location of the input and output
  buffers, so the buffers can't be deallocated, as it may be the case with the filter's input and output images.
\end{itemize}

The destruction of the input can be avoided in the creation plan by using the \verb$FFTW_ESTIMATE$ flag
-- that flag was already used in the implementation. It can also be avoided while running FFTW, for 
the real to complex transform, by also using the flag \verb$FFTW_PRESERVE_INPUT$. So in 
\verb$itk::FFTWRealToComplexConjugateImageFilter$, if the right flags are used, there is no need for
the input buffer.
The case is a bit
more difficult for the complex to real transform: FFTW does not provide any algorithm able to preserve
the input, so the input must be copied to an internal buffer. There is a case however, when we don't
care about destroying the input in ITK: when the \verb$ReleaseDataFlag$ is \verb$ON$ on the input of the filter.
In that case, the input will be deallocated right after the end of the execution of the filter, and so
there is no need to take care of the input -- it can be destroyed by the filter, and then deallocated.
\verb$itk::FFTWRealToComplexConjugateImageFilter$ has been modified to use the input image directly without 
intermediate buffer when \verb$ReleaseDataFlag$ is \verb$ON$. The detection of the \verb$ReleaseDataFlag$
is a bit difficult during the pipeline 
execution: this flag is always modified to \verb$OFF$ before the call to \verb$GenerateData()$ to avoid the 
destruction by a minipipeline. In consequence, the detection of the \verb$ReleaseDataFlag$ is done at an 
earlier stage in the pipeline execution, in \verb$UpdateOutputData()$, and stored in a member variable:
\verb$m_CanUseDestructiveAlgorithm$.

The later reuse of the plan, while sensible when not using the \verb$FFTW_ESTIMATE$ flag, is not much 
interesting when it is used: the time needed to create the plan is very short -- way shorter than the
FFT computation. Also, the number of threads and the image size are stored in the plan, and can't be 
changed later, which doesn't fit well with ITK pipeline model. In the enhanced version, the plan is
not kept between the filters execution, and so no intermediate output buffer are used, decreasing 
significantly the needed memory.

Another reason to create the internal buffer may be the need to align properly the image, to significantly
improve the FFTW performance by using the SIMD operations. The buffer were simply allocated with the
\verb$new$ operator, which is not enough to ensure a proper alignement
on some plateforms, including Microsoft Windows and Linux.

\subsubsection{Code safety}

Only \verb$fftwf_execute(plan);$ and \verb$fftw_execute(plan);$ are thread safe in FFTW -- the other functions
are not thread safe. In consequence, the call to all the other functions has been protected by a 
global lock in the enhanced classes.

Also, the check for the size of the image to know if the plan can be reused may not have been good 
enough, because it was testing only the number of pixels and not the size on all the dimensions. 
The plan is not kept between the execution, so this can not be a problem anymore.

@@

At the end of the convolution, the image is cropped to return to the size of the input image. Again
there is currently no option in ITK to do that fully in the pipeline model. In consequence, a new
filter has been developped for this task.

\verb$itk::RegionFromReferenceImageFilter$ is implemented as a subclass of \verb$itk::ExtractImageFilter$.
This filter requires two inputs -- the first is the image to crop, the second is simply used to get 
the region to extract in the first one. The reference image -- the second input -- is simply expected to
be of type \verb$itk::ImageBase$ with the same dimension than the first one, and so it's type don't
need to be specified in the template parameters.

@@

\verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
filters to make the FFT based convolution in a single and easy to use filter.
It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
user specified number of threads.

It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTZeroPaddingImageFilter$ to let
the user choose between the extra performance added by the padding and the extra memory cost.
